Infinite Cellular Automata Running Virtual 6502 Machines

### 2018-07-02 paper.tex
cube

The cellular automaton is a $L \times L \times L$ array of cells.
Cells are indexed $(x,y,z$) where $0 \leq x \leq L$, $0 \leq y \leq L$, $0 \leq z \leq L$.
Each cell has 256 bytes (one ``page'') of state.
All cells' bytes are initially zero.

@@

A cell $(x,y,z)$ is updated by emulating an 8-bit virtual machine.
The CPU is the MOS Technology 6502 processor.
The memory map is shown in \ref{tab:MemoryMap},
with detail for the memory-mapping of adjacent cells in \ref{tab:MooreNeighborhood}.

\begin{table}
\begin{tabular}{llll}
  \hline
  From & To & Type & Contents \\
  \hline
  {\tt 0000} & {\tt 00FF} & RAM (storage-mapped) & State of cell $(x,y,z)$ \\
  {\tt 0100} & {\tt 01FF} & RAM (transient) & Stack, initialized to zero \\
  {\tt 0200} & {\tt 03FF} & RAM (transient) & ``Scratch'' memory, initialized to zero \\
  {\tt 0400} & {\tt 0401} & ROM (emulator-mapped) & Cycles left before termination \\
  {\tt 0402} & {\tt 0402} & ROM (emulator-mapped) & Random byte generator \\
  {\tt 0400} & {\tt 3FFF} & ROM & Reserved, zero \\
  {\tt 4000} & {\tt 7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
  {\tt 8000} & {\tt FEFF} & ROM & Read-only (operating system) \\
  {\tt FF00} & {\tt FFFF} & ROM & Includes interrupt and reset vectors. All zero \\
  \hline
\end{tabular}
\caption{
  \label{tab:MemoryMap}
  Memory map for the 6502 virtual machine of cell $(x,y,z)$.
  Note that the BREAK vector {\tt FFFC-FFFD} is zero, so execution begins at {\tt 0000}.
}
\end{table}

\begin{table}
\mooremap
\caption{
  \label{tab:MooreNeighborhood}
  Memory map for the 6502 virtual machine of cell $(x,y,z)$,
  using the 26-cell three-dimensional Moore neighborhood.
}
\end{table}

All other bytes are read-only and zero; writing to them has no effect.
This can be understood as the cell $(x-1+i,y-1+j,z-1+k)$
mapping to page number (64+i*16+j*4+k) with $i,j,k \in \{0,1,2\}$.
Thus, in a memory-mapped 16-bit address, the bits have the following meanings

@@

>>>>

The cellular automaton is a $L \times L \times L$ array of cells.
Cells are indexed $(x,y,z$) where $0 \leq x \leq L$, $0 \leq y \leq L$, $0 \leq z \leq L$.

Each cell has 256 bytes (one ``page'') of state.
All cells' bytes are initially zero.
This is the ``storage'' of the virtual machine.

@@

A cell $(x,y,z)$ is updated by emulating an 8-bit virtual machine.
The CPU is the MOS Technology 6502 processor.
The memory map is shown in Table~\ref{tab:MemoryMap}.
In this memory map, the 27 pages of storage
for the cell and its 26 neighbors
in the $3 \times 3 \times 3$ unit cube centered on $(x,y,z)$
are mapped directly to pages in the RAM of the virtual machine.

The detailed description of the memory map for this $3 \times 3 \times 3$ neighborhood,
which in cellular automata terminology is
the {\em three-dimensional Moore neighborhood},
is given in Table~\ref{tab:MooreNeighborhood}.

The memory map can be understood as the cell $(x+i-1,y+j-1,z+k-1)$
mapping to page number (64+i*16+j*4+k) with $i,j,k \in \{0,1,2\}$.
Thus, in a memory-mapped 16-bit address, the bits have the following meanings

@@

\begin{table}
\begin{tabular}{llll}
  \hline
  From & To & Type & Contents \\
  \hline
  {\tt 0000} & {\tt 00FF} & RAM (storage-mapped) & State of cell $(x,y,z)$ \\
  {\tt 0100} & {\tt 01FF} & RAM (transient) & Stack, initialized to zero \\
  {\tt 0200} & {\tt 03FF} & RAM (transient) & ``Scratch'' memory, initialized to zero \\
  {\tt 0400} & {\tt 0401} & ROM (emulator-mapped) & Cycles left before termination \\
  {\tt 0402} & {\tt 0402} & ROM (emulator-mapped) & Random byte generator \\
  {\tt 0403} & {\tt 3FFF} & Reserved & Read-only, zero \\
  {\tt 4000} & {\tt 7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
  {\tt 8000} & {\tt FEFF} & ROM & Read-only (operating system) \\
  {\tt FF00} & {\tt FFFF} & ROM & Includes interrupt and reset vectors. All zero \\
  \hline
\end{tabular}
\caption{
  \label{tab:MemoryMap}
  Memory map for the 6502 virtual machine of cell $(x,y,z)$.
  Note that the BREAK vector {\tt FFFC-FFFD} is zero, so execution begins at {\tt 0000}.
}
\end{table}

\begin{table}
\mooremap
\caption{
  \label{tab:MooreNeighborhood}
  Memory map for the 6502 virtual machine of cell $(x,y,z)$,
  using the 26-cell three-dimensional Moore neighborhood.
  All address ranges not defined in this table are read-only and zero; writing to them has no effect.
}
\end{table}
